---
title: "Visualizing STRING-GSEA Results with enrichplot"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Visualizing STRING-GSEA Results with enrichplot}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6,
  dpi = 150
)
```

## Loading STRING-GSEA JSON results

The `read_gsea_json()` function reads a JSON file produced by the string-gsea
Python pipeline and returns a double-nested named list of `enrichResult` objects:
`results[["contrast"]][["category"]]`.

```{r load}
library(stringGSEAplot)
library(enrichplot)

json_path <- system.file("extdata", "WU2848501_gsea_result.json.gz",
                         package = "stringGSEAplot")
stopifnot(nzchar(json_path))

results <- read_gsea_json(json_path)

cat("Contrasts:", paste(names(results), collapse = ", "), "\n")
cat("Categories in first contrast:", paste(names(results[[1]]), collapse = ", "), "\n")
```

We'll pick one contrast and one category to demonstrate all plots.

```{r pick}
contrast_name <- names(results)[1]
er <- results[[contrast_name]][["GO Process"]]
er
```

## Dotplot

The classic enrichment dotplot. Size = gene count, color = adjusted p-value.

```{r dotplot, fig.height=8}
enrichplot::dotplot(er, showCategory = 20) +
  ggplot2::ggtitle(paste("GO Process -", contrast_name))
```

## Barplot

```{r barplot, fig.height=8}
barplot(er, showCategory = 20) +
  ggplot2::ggtitle(paste("GO Process -", contrast_name))
```

## Cnetplot (category-gene network)

Shows which genes are shared across enriched terms.

```{r cnetplot, fig.height=8, fig.width=10}
enrichplot::cnetplot(er, showCategory = 10,
  cex.params = list(category_label = 0.7, gene_label = 0.5))
```

## Heatplot

Gene-term heatmap showing which genes belong to which terms.

```{r heatplot, fig.height=6, fig.width=10}
enrichplot::heatplot(er, showCategory = 15)
```

## Enrichment map (emapplot)

Network of enriched terms connected by shared genes. Requires computing
pairwise term similarity first.

```{r emapplot, fig.height=8, fig.width=8}
er_sim <- enrichplot::pairwise_termsim(er)
enrichplot::emapplot(er_sim, showCategory = 30)
```

## Treeplot

Hierarchical clustering of enriched terms based on gene overlap.

```{r treeplot, fig.height=10, fig.width=10}
enrichplot::treeplot(er_sim, showCategory = 30)
```

## Upsetplot

UpSet-style plot showing intersections of gene sets.

```{r upsetplot, fig.height=6, fig.width=8}
enrichplot::upsetplot(er, n = 10)
```

## Comparing categories within one contrast

Let's compare a few categories side by side using dotplots.

```{r compare-categories, fig.height=12, fig.width=10}
categories_to_show <- c("GO Process", "KEGG", "Reactome", "WikiPathways")
available <- intersect(categories_to_show, names(results[[contrast_name]]))

plots <- lapply(available, function(cat) {
  enrichplot::dotplot(results[[contrast_name]][[cat]], showCategory = 10) +
    ggplot2::ggtitle(cat) +
    ggplot2::theme(plot.title = ggplot2::element_text(size = 10))
})

if (length(plots) > 0) {
  patchwork::wrap_plots(plots, ncol = 2)
}
```

## Cross-contrast comparisons with compareClusterResult

enrichplot supports comparing enrichment results across clusters (here:
contrasts) via the `compareClusterResult` class. `build_compareClusterResult()`
merges multiple `enrichResult` objects into one comparison object.

```{r compare-contrasts-setup}
# Build a compareClusterResult for GO Process across both contrasts
go_list <- lapply(results, function(x) x[["GO Process"]])
ccr <- build_compareClusterResult(go_list)
ccr
```

### Dotplot (cross-contrast)

Side-by-side dotplot comparing enriched GO terms across contrasts.

```{r compare-dotplot, fig.height=10, fig.width=10}
enrichplot::dotplot(ccr, showCategory = 10) +
  ggplot2::ggtitle("GO Process across contrasts")
```

### Cnetplot (cross-contrast)

Gene-category network showing shared genes across contrasts.

```{r compare-cnetplot, fig.height=8, fig.width=10}
enrichplot::cnetplot(ccr, showCategory = 5)
```

### Enrichment map (cross-contrast)

Network of enriched terms with pie charts showing contrast membership.

```{r compare-emapplot, fig.height=8, fig.width=10}
ccr_sim <- enrichplot::pairwise_termsim(ccr)
enrichplot::emapplot(ccr_sim, showCategory = 20)
```

### Comparing multiple categories across contrasts

```{r compare-multi-category, fig.height=16, fig.width=10}
categories_to_compare <- c("GO Process", "KEGG", "Reactome", "SMART")
available <- intersect(
  categories_to_compare,
  Reduce(intersect, lapply(results, names))
)

plots <- lapply(available, function(cat) {
  cat_list <- lapply(results, function(x) x[[cat]])
  ccr_cat <- build_compareClusterResult(cat_list)
  enrichplot::dotplot(ccr_cat, showCategory = 8) +
    ggplot2::ggtitle(cat) +
    ggplot2::theme(plot.title = ggplot2::element_text(size = 10))
})

if (length(plots) > 0) {
  patchwork::wrap_plots(plots, ncol = 1)
}
```

## Session info

```{r sessioninfo}
sessionInfo()
```
