---
title: "STRING-GSEA Enrichment Report"
metadata-files: ["_fgcz-report.yml"]
params:
  json_path: !expr system.file("extdata", "WU2848501_gsea_result.json.gz", package = "stringGSEAplot")
  show_category: 20
vignette: >
  %\VignetteIndexEntry{STRING-GSEA Enrichment Report}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup}
#| include: false
library(stringGSEAplot)

# Copy FGCZ report support files into the working directory
templates_dir <- system.file("templates", package = "stringGSEAplot")
for (f in c("_fgcz-report.yml", "fgcz_header_quarto.html")) {
  src <- file.path(templates_dir, f)
  if (nzchar(src) && file.exists(src) && !file.exists(f)) {
    file.copy(src, f)
  }
}

results <- read_gsea_json(params$json_path)
contrast_names <- names(results)
all_categories <- Reduce(union, lapply(results, names))
shared_categories <- Reduce(intersect, lapply(results, names))
run_metadata <- attr(results, "metadata")
run_links <- attr(results, "links")

plots_dir <- "plots"
dir.create(plots_dir, showWarnings = FALSE, recursive = TRUE)
plot_counter <- 0L

# Save a ggplot/grid object to PNG and return the relative path, or NULL on failure.
save_plot <- function(plot_obj, width = 10, height = 8, res = 150) {
  plot_counter <<- plot_counter + 1L
  fname <- sprintf("plot_%04d.png", plot_counter)
  fpath <- file.path(plots_dir, fname)
  tryCatch({
    grDevices::png(fpath, width = width, height = height, units = "in", res = res)
    on.exit(grDevices::dev.off())
    print(plot_obj)
    fpath
  }, error = function(e) {
    tryCatch(grDevices::dev.off(), error = function(e2) NULL)
    message("save_plot failed [", fname, "]: ", conditionMessage(e))
    NULL
  })
}

# Shortcut — single place to control default plot size
sp <- function(plot_obj) save_plot(plot_obj, width = 10, height = 8, res = 150)

# Try a plot expression, log any error and return NULL
try_plot <- function(label, expr) tryCatch(expr, error = function(e) { message(label, " failed: ", conditionMessage(e)); NULL })

# Number of significant terms in an enrichResult
n_terms <- function(er) nrow(er@result[er@result$p.adjust < 0.05, ])

# Truncate Description labels in an enrichResult (returns a copy)
truncate_descriptions <- function(er, max_chars = 60) {
  er_copy <- er
  desc <- er_copy@result$Description
  long <- nchar(desc) > max_chars
  desc[long] <- paste0(substr(desc[long], 1, max_chars - 3), "...")
  er_copy@result$Description <- desc
  er_copy
}

# Shortcut - adjust max_chars here to control label length globally
td <- function(er) truncate_descriptions(er, max_chars = 60)

# Columns to show in the summary DT table
table_cols <- c("ID", "Description", "GeneRatio", "BgRatio",
                "pvalue", "p.adjust", "qvalue", "Count")
```

::: {.panel-tabset}

# Settings

This report summarises gene set enrichment analysis (GSEA) results obtained via
the [STRING-DB](https://string-db.org/) API.
Pre-ranked gene lists were uploaded to STRING-DB, which performs enrichment
scoring based on the method described in
[Subramanian et al. (2005)](https://doi.org/10.1073/pnas.0506580102).
Enrichment categories include Gene Ontology, KEGG, Reactome, UniProt Keywords,
DISEASES, COMPARTMENTS, and others provided by STRING.
The interactive plots below are generated with
[enrichplot](https://bioconductor.org/packages/enrichplot/) and
[clusterProfiler](https://bioconductor.org/packages/clusterProfiler/).

```{r settings-metadata}
#| echo: false
if (!is.null(run_metadata)) {
  meta_df <- data.frame(
    Parameter = c("Workunit ID", "Species (NCBI Taxon)", "FDR threshold",
                   "Rank direction", "Caller identity", "STRING API"),
    Value = c(
      as.character(run_metadata$workunit_id),
      as.character(run_metadata$species),
      as.character(run_metadata$fdr),
      as.character(run_metadata$ge_enrichment_rank_direction),
      as.character(run_metadata$caller_identity),
      as.character(run_metadata$api_base_url)
    )
  )
  DT::datatable(
    meta_df, caption = "Analysis parameters", filter = "none",
    rownames = FALSE, class = "compact stripe",
    options = list(pageLength = 10, scrollX = TRUE, dom = "t")
  )
}
```

```{r settings-links}
#| echo: false
if (!is.null(run_links)) {
  links_df <- data.frame(
    Contrast = names(run_links),
    Link = vapply(run_links, function(url) {
      paste0("<a href='", url, "' target='_blank'>", url, "</a>")
    }, character(1)),
    stringsAsFactors = FALSE
  )
  DT::datatable(
    links_df, caption = "STRING-DB permanent links", filter = "none",
    rownames = FALSE, class = "compact stripe", escape = FALSE,
    options = list(pageLength = 20, scrollX = TRUE, dom = "t")
  )
}
```


```{r per-contrast}
#| echo: false
#| results: asis

for (cn in contrast_names) {
  cat("#", cn, "\n\n")
  categories <- names(results[[cn]])

  cat("::: {.panel-tabset}\n\n")

  for (cat_name in categories) {
    er <- results[[cn]][[cat_name]]
    nt <- n_terms(er)
    if (nt < 1) next

    cat("##", cat_name, "\n\n")
    cat("::: {.panel-tabset}\n\n")

    # --- Table -----------------------------------------------------------
    cat("### Table\n\n")
    top_df <- er@result[order(er@result$p.adjust), ]
    show_cols <- intersect(table_cols, colnames(top_df))
    top_df <- top_df[seq_len(min(params$show_category, nrow(top_df))), show_cols]
    tbl <- DT::datatable(
      top_df,
      caption = paste("Top enriched terms -", cat_name),
      filter = "none", rownames = FALSE, class = "compact stripe",
      options = list(pageLength = 20, scrollX = TRUE, dom = "tip")
    )
    # DT widgets need special handling in results:asis chunks
    tbl_html <- as.character(htmltools::tagList(tbl))
    cat(tbl_html)
    cat("\n\n")

    er_td <- td(er)
    n_show <- min(params$show_category, nt)

    # --- Dotplot ---------------------------------------------------------
    p <- try_plot("dotplot", enrichplot::dotplot(er_td, showCategory = n_show) +
        ggplot2::ggtitle(paste(cat_name, "-", cn)))
    if (!is.null(p)) {
      fpath <- sp(p)
      if (!is.null(fpath)) {
        cat("### Dotplot\n\n")
        cat(sprintf("![Dotplot: dot size = gene count, color = adjusted p-value. Top %d enriched %s terms for %s.](%s){width=600px}\n\n", n_show, cat_name, cn, fpath))
      }
    }

    # --- Barplot ---------------------------------------------------------
    p <- try_plot("barplot", barplot(er_td, showCategory = n_show) +
        ggplot2::ggtitle(paste(cat_name, "-", cn)))
    if (!is.null(p)) {
      fpath <- sp(p)
      if (!is.null(fpath)) {
        cat("### Barplot\n\n")
        cat(sprintf("![Barplot: bar length = gene count, color = adjusted p-value. Top %d enriched %s terms for %s.](%s){width=600px}\n\n", n_show, cat_name, cn, fpath))
      }
    }

    # --- Cnetplot --------------------------------------------------------
    if (nt >= 2) {
      n_cnet <- min(10, nt)
      p <- try_plot("cnetplot", enrichplot::cnetplot(er_td, showCategory = n_cnet,
          cex.params = list(category_label = 0.7, gene_label = 0.5)))
      if (!is.null(p)) {
        fpath <- sp(p)
        if (!is.null(fpath)) {
          cat("### Cnetplot\n\n")
          cat(sprintf("![Concept network: links between enriched terms and their associated genes, showing which genes are shared across %s categories.](%s){width=700px}\n\n", cat_name, fpath))
        }
      }
    }

    # --- Heatplot --------------------------------------------------------
    if (nt >= 2) {
      n_heat <- min(15, nt)
      p <- try_plot("heatplot", enrichplot::heatplot(er_td, showCategory = n_heat))
      if (!is.null(p)) {
        fpath <- sp(p)
        if (!is.null(fpath)) {
          cat("### Heatplot\n\n")
          cat(sprintf("![Heatplot: gene-term association matrix. Color intensity = fold change. Rows = top %d %s terms, columns = genes.](%s){width=700px}\n\n", n_heat, cat_name, fpath))
        }
      }
    }

    # --- Enrichment Map --------------------------------------------------
    if (nt >= 3) {
      n_emap <- min(30, nt)
      p <- try_plot("emapplot", {
        er_sim <- enrichplot::pairwise_termsim(er_td)
        enrichplot::emapplot(er_sim, showCategory = n_emap)
      })
      if (!is.null(p)) {
        fpath <- sp(p)
        if (!is.null(fpath)) {
          cat("### Enrichment Map\n\n")
          cat(sprintf("![Enrichment map: network of %s terms connected by gene overlap similarity. Node size = gene count, color = p-value. Clustered terms share biological function.](%s){width=600px}\n\n", cat_name, fpath))
        }
      }
    }

    # --- Treeplot --------------------------------------------------------
    if (nt >= 3) {
      n_tree <- min(30, nt)
      p <- try_plot("treeplot", {
        er_sim <- enrichplot::pairwise_termsim(er_td)
        enrichplot::treeplot(er_sim, showCategory = n_tree)
      })
      if (!is.null(p)) {
        fpath <- sp(p)
        if (!is.null(fpath)) {
          cat("### Treeplot\n\n")
          cat(sprintf("![Treeplot: hierarchical clustering of %s terms by semantic similarity. Branches group functionally related terms; bar length = gene count.](%s){width=700px}\n\n", cat_name, fpath))
        }
      }
    }

    # --- UpSet Plot ------------------------------------------------------
    if (nt >= 2) {
      n_upset <- min(10, nt)
      p <- try_plot("upsetplot", enrichplot::upsetplot(er_td, n = n_upset))
      if (!is.null(p)) {
        fpath <- sp(p)
        if (!is.null(fpath)) {
          cat("### UpSet Plot\n\n")
          cat(sprintf("![UpSet plot: set intersections of genes across top %d %s terms. Dots and lines show which term combinations share genes; bars = intersection size.](%s){width=600px}\n\n", n_upset, cat_name, fpath))
        }
      }
    }

    # --- Ridgeplot -------------------------------------------------------
    if (nt >= 2) {
      n_ridge <- min(params$show_category, nt)
      p <- try_plot("ridgeplot", {
        raw <- attr(results, "raw_data")[[cn]]
        gr  <- build_gseaResult(raw$cats[[cat_name]], raw$gene_pool, raw$rank_list)
        enrichplot::ridgeplot(gr, showCategory = as.numeric(n_ridge)) +
          ggplot2::ggtitle(paste(cat_name, "-", cn))
      })
      if (!is.null(p)) {
        fpath <- sp(p)
        if (!is.null(fpath)) {
          cat("### Ridgeplot\n\n")
          cat(sprintf("![Ridgeplot: distribution of gene ranking scores for genes in each enriched %s term. Color = adjusted p-value.](%s){width=700px}\n\n", cat_name, fpath))
        }
      }
    }

    cat(":::\n\n")
  }

  cat(":::\n\n")
}
```


# Cross-Contrast

::: {.callout-note collapse="true"}
## About
Comparison of enrichment results across all contrasts using
`compareClusterResult` objects from `build_compareClusterResult()`.
:::

```{r cross-contrast}
#| echo: false
#| results: asis

cat("::: {.panel-tabset}\n\n")

# --- Dotplot -----------------------------------------------------------
plots <- list()
for (cat_name in shared_categories) {
  cat_list <- lapply(results, function(x) td(x[[cat_name]]))
  p <- try_plot("cross dotplot", {
    ccr <- build_compareClusterResult(cat_list)
    enrichplot::dotplot(ccr, showCategory = 10) +
      ggplot2::ggtitle(paste(cat_name, "- across contrasts"))
  })
  if (!is.null(p)) plots[[cat_name]] <- p
}
if (length(plots) > 0) {
  cat("## Dotplot\n\n")
  cat("::: {.panel-tabset}\n\n")
  for (cat_name in names(plots)) {
    fpath <- sp(plots[[cat_name]])
    if (!is.null(fpath)) {
      cat("###", cat_name, "\n\n")
      cat(sprintf("![Cross-contrast dotplot: comparing enrichment of %s terms across all contrasts. Dot size = gene count, color = adjusted p-value.](%s){width=700px}\n\n", cat_name, fpath))
    }
  }
  cat(":::\n\n")
}

# --- Cnetplot ----------------------------------------------------------
plots <- list()
for (cat_name in shared_categories) {
  cat_list <- lapply(results, function(x) td(x[[cat_name]]))
  p <- try_plot("cross cnetplot", {
    ccr <- build_compareClusterResult(cat_list)
    enrichplot::cnetplot(ccr, showCategory = 5)
  })
  if (!is.null(p)) plots[[cat_name]] <- p
}
if (length(plots) > 0) {
  cat("## Cnetplot\n\n")
  cat("::: {.panel-tabset}\n\n")
  for (cat_name in names(plots)) {
    fpath <- sp(plots[[cat_name]])
    if (!is.null(fpath)) {
      cat("###", cat_name, "\n\n")
      cat(sprintf("![Cross-contrast concept network: %s terms and their associated genes across all contrasts, showing shared gene membership.](%s){width=700px}\n\n", cat_name, fpath))
    }
  }
  cat(":::\n\n")
}

# --- Enrichment Map ----------------------------------------------------
plots <- list()
for (cat_name in shared_categories) {
  cat_list <- lapply(results, function(x) td(x[[cat_name]]))
  p <- try_plot("cross emapplot", {
    ccr <- build_compareClusterResult(cat_list)
    ccr_sim <- enrichplot::pairwise_termsim(ccr)
    enrichplot::emapplot(ccr_sim, showCategory = 20)
  })
  if (!is.null(p)) plots[[cat_name]] <- p
}
if (length(plots) > 0) {
  cat("## Enrichment Map\n\n")
  cat("::: {.panel-tabset}\n\n")
  for (cat_name in names(plots)) {
    fpath <- sp(plots[[cat_name]])
    if (!is.null(fpath)) {
      cat("###", cat_name, "\n\n")
      cat(sprintf("![Cross-contrast enrichment map: network of %s terms across all contrasts, connected by gene overlap similarity.](%s){width=700px}\n\n", cat_name, fpath))
    }
  }
  cat(":::\n\n")
}

cat(":::\n\n")
```


# Session Info

```{r session-info}
#| echo: false
utils::sessionInfo()
```

:::
